---
title: "Working with Waveform Data using fluoR"
author: "Andrew Tamalunas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Waveform Data using fluoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(fluoR)
```

Calcium indicator methods such as GCaMP produce massive amounts of data; in some cases producing hundreds-of-thousands of data points for a single subject. Further, there is currently no ubiquitous way to organize or analyze this type of data. To even analyze the data, the researcher must:

* Organize the data into a format that is easy to manipulate

* Visualize and explore the data 

* Extract useful data points for further visualization and analysis

The fluoR package gets researchers to the analysis phase more quickly by:

* Providing simple, quick commands to format data for use both with, and without the fluoR package

* Including shortcuts for data exploration and useful visualizations

* Extracting data points and waveform characteristics for summarizing and comparing activity within and between trials

This document explains how to use fluoR's functions to format, extract, and manipulate calcium indicator data. 

# Data: GCaMP

The data included with the fluoR provides a sample of a time series-like dataset exported from Matlab using the [TDTFilter command](https://www.tdt.com/support/sdk.html) with a modified version of [Dr. David Root's Matlab script](https://www.tdt.com/support/EXEpocAveragingExampleDR.html). This data was collected using GCaMP6.

This dataset consists of 11 rows and 814 columns. 10 trials from a pilot study were used, with calcium activity from 4 seconds before and after stimulus onset (0s). 

# Data formatting

In order to use the rest of the package, data must be in a data frame format, with the requirements being:

1. The first column is a measure of time

2. The following columns are recorded values from trials in ascending order that correspond to the times the values were recorded at, with 1 column per trial

Fortunately, the fluoR package includes functions that quickly reformat the data for ease with both user manipulation and use of this package. All formatting commands output this type of data frame.

### Format organized data with format_data

Currently, the only command for formatting data is `format_data`, which takes a matrix of a time measurement in the first column or row, and one trial per column or row in the subsequent columns/rows. A data frame with the first row "Time" and subsequent rows "Trial#" is outputted.

```{r}
df <- format_data(GCaMP)

### Old table
GCaMP[1:5, 1:5]

### New table
df[1:5, 1:5]
```

Note: the data frame used with the GCalcium package does not have to be labeled the same as the format_data frame. This is simply for ease of calling each trial using outside functions.

# Extracting useful information for analysis

To perform analyses or explore differences in activity waveforms, one must filter and summarize the data. Knowing what wave characteristics to compare can be confusing; as many scientists do not typically work with this type of data. The following commands extract and/or summarize numbers comparisons that have been used in past research. These functions are split into 2 types: vector inputs, and matrix (`format_data` style) inputs.

## Vector 

* `find_peaks` looks for peaks based on inflection points and a simple filter

* `inflect_points` finds the peaks and valleys of waves

* `percent_change` transforms a numeric vector into percent change from a base value

* `z_score` calculates a standard or modified z-score 

### find_peaks

`find_peaks` finds peaks or valleys in waveforms by using inflection points, with filter of 'n' increasing/decreasing points on both sides of each inflection point. A positive numerical input for 'n.points' returns the indices of peaks for the input vector, while a negative value returns indices of valleys.

Let's say we wanted to find all peaks of trial 1 that have 10 consecutive decreasing points on the left and right of each peak, and use these indices to subset the data.

```{r}
peak.indices <- find_peaks(xvals = df$Trial1, 
                           n.points = 10)

## Subset using indexing
peaks.df <- data.frame(
  times = df$Time[peak.indices],
  vals = df$Trial1[peak.indices]
)

head(peaks.df, 4)
```

### inflect_points

`inflect_points` uses derivatives to find and label the inflection points (peaks and valleys) of a vector, along with the points between them. 

```{r}
inf.pts <- inflect_points(x = df$Trial1)

print(inf.pts[1:45])
```

The value -2 indicates a peak, 2 indicates a valley, and 0 indicates a point on the curve between -2 and 2, or vice versa. 

### percent_change

`percent_change` is a simple function that calculates the percent change from a base value.

If we wanted to transform trial 1's values to percent change from the average baseline value for the first trial, we'd enter

```{r}
baseline.mean <- mean(df$Trial1[df$Time > -4 & df$Time < 0])

perc.change <- percent_change(xvals = df$Trial1,
                              base.val = baseline.mean)

print(perc.change[1:45])
```

### z_score

`z_score` is a simple function that transforms a numeric vector to standard or modified z-score values.

Percent change from baseline is not always an effective metric. It normalizes your values, but does not standardize and is very susceptible to outliers.

Let's say we wanted to transform trial 1's values into z-scores, in relation to baseline. This would be done to take the baseline deviance (where there is presumably less activity) into consideration and help reveal significant changes in intensity.

```{r}
baseline.mean <- mean(df$Trial1[df$Time > -4 & df$Time < 0])
baseline.sd <- sd(df$Trial1[df$Time > -4 & df$Time < 0])

z.scores <- z_score(xvals = df$Trial1,
        mu = baseline.mean,
        sigma = baseline.sd)

print(z.scores[1:45])
```

Standard z-scores, however, are not always the optimal transformation for fluorescence waveform data. This type of data often includes spikes (outliers), which can cause the mean to be a subpar measure.

To solve this, we can use a median-based modified z-score.

```{r}
baseline.median <- median(df$Trial1[df$Time > -4 & df$Time < 0])
baseline.mad <- mad(df$Trial1[df$Time > -4 & df$Time < 0])

mod.z.scores <- z_score(xvals = df$Trial1,
        mu = baseline.median,
        sigma = baseline.mad,
        z.type = 'modified')

print(mod.z.scores[1:45])
```

## Matrix or Data frame

* `baseline_transform` is a dataframe-based wrapper for the `z_score` and `percent_baseline` commands

* `inflect_points_df` finds provides a summary of the data between each inflection point (half wave)

* `plot_trials` creates a simple plot for a set of trials

* `summarize_trials` calculates a summary statistic; collapsing across trials for each timestamp

### baseline_transform

`baseline_transform` lets the user utilize the `percent_change` and `z_score` commands without isolating the values from a subject's data frame. It also speeds up the process of analyzing multiple trials.

To transform the values from trials 1-5 into percent change, we'd use

```{r}
t1thru5 <- baseline_transform(dataframe = df,
                   trials = 1:5,
                   baseline.times = c(-4, 4),
                   type = 'percent_change')
print(head(t1thru5))
```

Alternatively, we can use `type = z_standard` or `type = z_modified`

### inflect_trials_df

`inflect_trials_df` is a dataframe-based version of `inflect_trials`. This 

### plot_trials

`plot_trials` uses base R graphics to create a quick plot of the trial waves. This is particularly useful when manually looking through the data and comparing trials.

Let's pretend that we introduced a stimulus at the end of trial 6. If we wanted to quickly look at how the subject's behavior was influenced in subsequent trials, we can simply enter

```{r, plot_trials, fig.height = 4, fig.width = 7}
plot_trials(dataframe = df,
            trials = 7:10)
```

### summarize_trials

`summarize_trials` lets the user specify a type of summary statistic (e.g. mean, standard deviation) to compute across a range of trials

For example: let's say we wanted to plot the mean for trials 1-5 and compare them to the mean of trials 6-10.

```{r, summarize_trials, fig.height = 4, fig.width = 7}
### Find means values across trials 1-5 and 6-10
summ1 <- summarize_trials(dataframe = df,
                          trials = 1:5,
                          summary.type = 'mean')
summ2 <- summarize_trials(df, 6:10, 'mean')

### Compare trial ranges
plot(x = summ1$Time,
     y = summ1$summ.trials,
     type = 'l',
     xlim = c(-4, 4),
     ylim = c(min(c(summ1[,2], summ2[,2])),
              max(c(summ1[,2], summ2[,2]))),
     xlab = 'Time',
     ylab = 'Intensity',
     col = 'blue') # plot using trials 1-5, using min/max of all trials for y limits

lines(x = summ2$Time,
      y = summ2$summ.trials,
      col = 'red') # add lines for 6-10

legend(x = 'topright', 
       legend = c('Trials 1-5', 'Trials 6-10'),
       col = c('blue', 'red'),
       lty = 1,
       bty = 'o',
       box.lwd = .01,
       cex = 0.75)
```